The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-02-15 script/en/chapters/digitalsignatures.tex
Corrected typos and overfull hbox; congruent index (no redundancy in wording; adjectives in lowercase in German; all normal words in lowercase in English)

\newpage
% --------------------------------------------------------------------------
\hypertarget{Chapter_Hashes-and-Digital-Signatures}{}
\chapter{Hash Functions and Digital Signatures}
\label{Chapter_Hashes-and-Digital-Signatures}
\index{Signature!digital}
(Schneider J. / Esslinger B. / Koy H., June 2002;
Updates: Feb. 2003, June 2005, July 2009, Nov. 2012)
% --------------------------------------------------------------------------
%\vskip + 30pt
\begin{ctsquote}
We can make everything out of this world, but we cannot create a world,
where humans in some ten thousand years can think: 'Ok, now it is enough.
Everything should stay like it is. Let's do no changes any more, don't do
inventions any more, because it cannot become better, and if, then we
don't want this.'
\caption[Stanislaw Lem]{Stanislaw Lem\footnotemark}\index{Lem, Stanislaw}
\end{ctsquote}
\addtocounter{footnote}{0}\footnotetext{This was the answer of Stanislaw Lem
to heavy critics at his philosophical main book ~``Summa Technologiae'',
1964, where he thought about the possibility of an evolution creating
artificial intelligence.}

%\vspace{12pt}
\noindent The aim of digital signatures is to guarantee the following two points:
\begin{itemize}
 \item User authenticity: \index{Authenticity!user} \\
      It can be checked whether a message really does
come from a particular person.
 \item Message integrity:  \index{Message integrity} \\
      It can be checked whether the message has been
changed (on route).
\end{itemize}

@@

In detail, a \index{Signature procedure} {\em Signature procedure}\footnote{%
With CT1\index{CrypTool 1} you can also generate and check 
digital signatures: Using\\
the submenus of the main menu {\bf Digital Signatures / PKI}  or using\\
menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
\textbackslash{} Signature Demonstration (Signature Generation)}.\\
Also with JCT\index{JCrypTool} (in the default and the algorithm perspektive)
its possible to create different kinds of electronic signatures.
} looks like
this: \\ Senders use their message and secret key to calculate the digital
signature for the message. Compared to hand-written signatures, digital
signatures therefore have the advantage that they also depend on the document to
be signed. Signatures from one and the same participant are different unless the
signed documents are completely identical. Even inserting a blank in the text
would lead to a different signature. The recipient of the message would
therefore detect any injury to the message integrity as this would mean that the
signature no longer matches the document and is shown to be incorrect when
verified.

The document is sent to the recipient together with the signature. The recipient
can then use the sender's public key, the document and the signature to
establish whether or not the signature is correct. The procedure we just described
has in practice, however, a decisive disadvantage. The signature would be
approximately as long as the document itself. To prevent an unnecessary increase
in data traffic, and also for reasons of performance\index{Performance}, we apply
a cryptographic hash function\footnote{%
Hash functions\index{Hash function} are implemented within 
CT1\index{CrypTool 1} at several places.\\
Using menus {\bf Individual Procedures \textbackslash{} Hash} and
            {\bf Analysis \textbackslash{} Hash}
you have the possibilities
% hier die items nicht einruecken!
\begin{list}{\textbullet}{\leftmargin10pt\addtolength{\itemsep}{-1.0\baselineskip}}
%\begin{itemize}\addtolength{\itemsep}{-1.0\baselineskip} % um den Abstand zu verringern bebe
\item to apply one of 6 hash functions to the content of the current window, \\
\item to calculate the hash value of a file, \\
\item to test, how changes to a text change the according hash value,\\
\item to calculate a key from a password according to the PKCS\#5 
      standard\index{PKCS\#5}, \\
\item to calculate HMACs from a text and a secret key, and\\
\item to perform a simulation, how digital signatures could be attacked 
      by a targeted search for hash value collisions\index{Collision}.
\end{list}
CT2\index{CrypTool 2} and JCT\index{JCrypTool} also contain different hash
methods: See the functions' lists within the appendix
\ref{s:appendix-template-overview-CT2} and
\ref{s:appendix-function-overview-JCT}.}
to the document -- before signing. The output of the hash function will 
then be signed.

% --------------------------------------------------------------------------
\hypertarget{Hash-functions-ht}{}
\section{Hash functions}
\index{Hash function}
A {\em hash function}\footnote{%
Hash algorithms compute a condensed representation of electronic data
(message).
When a message is input to a hash algorithm, the result is an output called
a message digest. The message digests typically range in length 
from 128 to 512 bits, depending on the algorithm. 
Secure hash algorithms are typically used with
other cryptographic algorithms, such as digital signature algorithms and
keyed-hash message authentication codes, or in the generation of random
numbers (bits).
}
maps a message of any length to a string of characters with a constant size,
the \index{Hash value} hash value.

@@

Cryptographically secure hash functions fulfill the following three requirements
(the order is in a way that the requirements increase):
\begin{itemize}
 \item Resistance against 1st pre-image attacks:
      \index{Pre-image attack!1st} \index{Attack!pre-image!1st}   \\
      It should be practically impossible, for a given number, to find a
      message that has precisely this number as hash value. \\
      Given (fix): hash value H', \\
      Searched: message m, so that: H(m) = H'.
 \item Resistance against 2nd pre-image attacks:
      \index{Pre-image attack!2nd} \index{Attack!pre-image!2nd}   \\
      It should be practically impossible, for a given message, to find another
      message, which has precisely the same hash value. \\
      Given (fix): message m1 [and so the hash value H1 = H(m1)], \\
      Searched: message m2, so that: H(m2) = H1.
 \item Collision resistance:
      \index{Collision resistance}  \\
      It should be practically impossible to find any two messages with the
      same hash value (it doesn't matter what hash value). \\
      Searched: 2 messages m1 and m2, so that: H(m1) = H(m2).
\end{itemize}

@@

  The output length of the SHA algorithms was enhanced because of the
  possibility of birthday attacks\index{Attack!birthday}\index{Collision}:
  these make n-bit AES and a 2n-bit hash roughly equivalent: \\
  - 128-bit AES -- SHA-256 \\
  - 192-bit AES -- SHA-384 \\
  - 256-bit AES -- SHA-512.

  With CT1\index{CrypTool 1} you can comprehend the birthday attack
  \index{Attack!birthday} on digital signatures: \\
  using the menu {\bf Analysis \textbackslash{} Hash \textbackslash{} 
  Attack on the Hash Value of the Digital Signature}.\\
  CT2\index{CrypTool 2} contains an MD5 collider.
  }  
or \index{RIPEMD-160} RIPEMD-160\footnote{%
  RIPEMD-160, RIPEMD-128 and the optional extension RIPEMD-256 have object
  identifiers defined by the ISO-identified organization TeleTrusT, both 
  as hash algorithm and in combination with RSA.
  RIPEMD-160 is also part of the ISO/IEC international standard 
  ISO/IEC 10118-3:1998 on dedicated hash functions, together with 
  RIPEMD-128 and SHA-1.
  Further details: \\
- \url{http://www.esat.kuleuven.ac.be/~bosselae/ripemd160.html}\\
- \url{http://www.ietf.org/rfc/rfc2857.txt} (``The Use of HMAC-RIPEMD-160-96
   within ESP and AH'').
  }%
) were trusted.

@@

% --------------------------------------------------------------------------
\vskip + 15pt
\section{RSA signatures}
\index{Signature!digital}
\index{RSA!signature}

@@

% --------------------------------------------------------------------------
\vskip + 15pt
\section{DSA signatures}
\index{Signature!digital}
\index{DSA!signature}

@@

The algorithm is a variant of the ElGamal scheme. Its security is based on
the Discrete Logarithm Problem\index{Logarithm problem!discrete}. The DSA public and private key and its
procedures for signature and verification are summarised in crypto procedure~\ref{dsasigproc}.

@@

\paragraph*{Signing}\strut\\
\begin{tabular}{l@{ }l}
$m$ & the message to be signed\\
$k$ & choose at random\index{Random}, less than $q$\\
$r$ & $= (g^k \; \mbox{ mod } p) \mbox{ mod } q$\\
$s$ & $= (k^{-1}(\mbox{SHA-1}(m) + xr)) \mbox{ mod } q$
\end{tabular}

\emph{Remark:}
\begin{itemize}
\item $(s,r)$ is the signature.
\item The security of the signature depends not only on the mathematical
  properties, but also on using a good random source\index{Random}  for $k$.
\item SHA-1 \index{SHA-1} is a 160-bit hash function.
\end{itemize}
\paragraph*{Verifying}\strut\\
\begin{tabular}{l@{ }l}
$w$ & $= s^{-1} \;  \mbox{ mod } q$\\
$u_1$ & $= (\mbox{SHA-1}(m)w) \mbox{ mod } q$\\
$u_2$ & $= (rw)  \mbox{ mod } q$\\
$v$ & $= (g^{u_1}y^{u_2}) \mbox{ mod } p)  \mbox{ mod } q$\\

@@

% --------------------------------------------------------------------------
\vskip + 15pt
\section{Public key certification}
\index{Certification!public key}
\index{PKI}
The aim of public key certification is to guarantee the connection between a
public key and a user and to make it traceable for external parties. In cases in
which it is impossible to ensure that a public key really belongs to a
particular person, many protocols are no longer secure, even if the individual
cryptographic modules cannot be broken.

% --------------------------------------------------------------------------
\vskip + 15pt
\subsection{Impersonation attacks}\label{Impersonalisierungsattacke}
\index{Impersonation attack}
Assume Charlie has two pairs of keys (PK1, SK1) and (PK2, SK2), where SK denotes
the secret key and PK the public key. Further assume that he manages to palm off
PK1 on Alice as Bob's public key and PK2 on Bob as Alice's public key (by
falsifying a public key directory).

@@

Another name for this type of attack is \index{Attack!man-in-the-middle}
``man-in-the-middle attack''. Users are promised
protection against this type of attack by public-key
certification, which is intended to guarantee the
\index{Authenticity} authenticity of public keys. The most common
certification method is the X.509 standard.

% --------------------------------------------------------------------------
\vskip + 15pt
\subsection{X.509 certificate}
\index{X.509} 
Each participant who wants to have an X.509 certificate (\cite{ds:X.509})
verifying that his public key belongs to a real person consults what is
known as a \index{Certification authority (CA)} certification authority
(CA)\footnote{%
Often called trust center, if the certificates are not only offered
to a closed user group.
}. 
He proves his identity to this CA (for example by showing his ID). 
The CA then issues him an electronic document (certificate) which 
essentially contains the name of the certificate-holder and the name
of the CA, the certificate-holder's public key and the validity period
of the certificate. The CA then signs the certificate using its secret key.

>>>>

\newpage
% --------------------------------------------------------------------------
\hypertarget{Chapter_Hashes-and-Digital-Signatures}{}
\chapter{Hash Functions and Digital Signatures}
\label{Chapter_Hashes-and-Digital-Signatures}
\index{signature!digital}
(Schneider J. / Esslinger B. / Koy H., June 2002;
Updates: Feb. 2003, June 2005, July 2009, Nov. 2012)
% --------------------------------------------------------------------------
%\vskip + 30pt
\begin{ctsquote}
We can make everything out of this world, but we cannot create a world,
where humans in some ten thousand years can think: 'Ok, now it is enough.
Everything should stay like it is. Let's do no changes any more, don't do
inventions any more, because it cannot become better, and if, then we
don't want this.'
\caption[Stanislaw Lem]{Stanislaw Lem\footnotemark}\index{Lem, Stanislaw}
\end{ctsquote}
\addtocounter{footnote}{0}\footnotetext{This was the answer of Stanislaw Lem
to heavy critics at his philosophical main book ~``Summa Technologiae'',
1964, where he thought about the possibility of an evolution creating
artificial intelligence.}

%\vspace{12pt}
\noindent The aim of digital signatures is to guarantee the following two points:
\begin{itemize}
 \item User authenticity: \index{authenticity!user} \\
      It can be checked whether a message really does
come from a particular person.
 \item Message integrity:  \index{message integrity} \\
      It can be checked whether the message has been
changed (on route).
\end{itemize}

@@

In detail, a \index{signature procedure} {\em Signature procedure}\footnote{%
With CT1\index{CrypTool 1} you can also generate and check 
digital signatures: Using\\
the submenus of the main menu {\bf Digital Signatures / PKI}  or using\\
menu {\bf Indiv. Procedures \textbackslash{} RSA Cryptosystem 
\textbackslash{} Signature Demonstration (Signature Generation)}.\\
Also with JCT\index{JCrypTool} (in the default and the algorithm perspektive)
its possible to create different kinds of electronic signatures.
} looks like
this: \\ Senders use their message and secret key to calculate the digital
signature for the message. Compared to hand-written signatures, digital
signatures therefore have the advantage that they also depend on the document to
be signed. Signatures from one and the same participant are different unless the
signed documents are completely identical. Even inserting a blank in the text
would lead to a different signature. The recipient of the message would
therefore detect any injury to the message integrity as this would mean that the
signature no longer matches the document and is shown to be incorrect when
verified.

The document is sent to the recipient together with the signature. The recipient
can then use the sender's public key, the document and the signature to
establish whether or not the signature is correct. The procedure we just described
has in practice, however, a decisive disadvantage. The signature would be
approximately as long as the document itself. To prevent an unnecessary increase
in data traffic, and also for reasons of performance\index{performance}, we apply
a cryptographic hash function\footnote{%
Hash functions\index{hash function} are implemented within 
CT1\index{CrypTool 1} at several places.\\
Using menus {\bf Individual Procedures \textbackslash{} Hash} and
            {\bf Analysis \textbackslash{} Hash}
you have the possibilities
% hier die items nicht einruecken!
\begin{list}{\textbullet}{\leftmargin10pt\addtolength{\itemsep}{-1.0\baselineskip}}
%\begin{itemize}\addtolength{\itemsep}{-1.0\baselineskip} % um den Abstand zu verringern bebe
\item to apply one of 6 hash functions to the content of the current window, \\
\item to calculate the hash value of a file, \\
\item to test, how changes to a text change the according hash value,\\
\item to calculate a key from a password according to the PKCS\#5 
      standard\index{PKCS\#5}, \\
\item to calculate HMACs from a text and a secret key, and\\
\item to perform a simulation, how digital signatures could be attacked 
      by a targeted search for hash value collisions\index{collision}.
\end{list}
CT2\index{CrypTool 2} and JCT\index{JCrypTool} also contain different hash
methods: See the functions' lists within the appendix
\ref{s:appendix-template-overview-CT2} and
\ref{s:appendix-function-overview-JCT}.}
to the document -- before signing. The output of the hash function will 
then be signed.

% --------------------------------------------------------------------------
\hypertarget{Hash-functions-ht}{}
\section{Hash functions}
\index{hash function}
A {\em hash function}\footnote{%
Hash algorithms compute a condensed representation of electronic data
(message).
When a message is input to a hash algorithm, the result is an output called
a message digest. The message digests typically range in length 
from 128 to 512 bits, depending on the algorithm. 
Secure hash algorithms are typically used with
other cryptographic algorithms, such as digital signature algorithms and
keyed-hash message authentication codes, or in the generation of random
numbers (bits).
}
maps a message of any length to a string of characters with a constant size,
the \index{hash value} hash value.

@@

Cryptographically secure hash functions fulfill the following three requirements
(the order is in a way that the requirements increase):
\begin{itemize}
 \item Resistance against 1st pre-image attacks:
      \index{pre-image attack!1st} \index{attack!pre-image!1st}   \\
      It should be practically impossible, for a given number, to find a
      message that has precisely this number as hash value. \\
      Given (fix): hash value H', \\
      Searched: message m, so that: H(m) = H'.
 \item Resistance against 2nd pre-image attacks:
      \index{pre-image attack!2nd} \index{attack!pre-image!2nd}   \\
      It should be practically impossible, for a given message, to find another
      message, which has precisely the same hash value. \\
      Given (fix): message m1 [and so the hash value H1 = H(m1)], \\
      Searched: message m2, so that: H(m2) = H1.
 \item Collision resistance:
      \index{collision resistance}  \\
      It should be practically impossible to find any two messages with the
      same hash value (it doesn't matter what hash value). \\
      Searched: 2 messages m1 and m2, so that: H(m1) = H(m2).
\end{itemize}

@@

  The output length of the SHA algorithms was enhanced because of the
  possibility of birthday attacks\index{attack!birthday}\index{collision}:
  these make n-bit AES and a 2n-bit hash roughly equivalent: \\
  - 128-bit AES -- SHA-256 \\
  - 192-bit AES -- SHA-384 \\
  - 256-bit AES -- SHA-512.

  With CT1\index{CrypTool 1} you can comprehend the birthday attack
  \index{attack!birthday} on digital signatures: \\
  using the menu {\bf Analysis \textbackslash{} Hash \textbackslash{} 
  Attack on the Hash Value of the Digital Signature}.\\
  CT2\index{CrypTool 2} contains an MD5 collider.
  }  
or \index{RIPEMD-160} RIPEMD-160\footnote{%
  RIPEMD-160, RIPEMD-128 and the optional extension RIPEMD-256 have object
  identifiers defined by the ISO-identified organization TeleTrusT, both 
  as hash algorithm and in combination with RSA.
  RIPEMD-160 is also part of the ISO/IEC international standard 
  ISO/IEC 10118-3:1998 on dedicated hash functions, together with 
  RIPEMD-128 and SHA-1.
  Further details: \\
- \url{http://www.esat.kuleuven.ac.be/~bosselae/ripemd160.html}\\
- \url{http://www.ietf.org/rfc/rfc2857.txt} (``The Use of HMAC-RIPEMD-160-96
   within ESP and AH'').
  }%
) were trusted.

@@

% --------------------------------------------------------------------------
\vskip + 15pt
\section{RSA signatures}
\index{signature!digital}
\index{RSA!signature}

@@

% --------------------------------------------------------------------------
\vskip + 15pt
\section{DSA signatures}
\index{signature!digital}
\index{DSA!signature}

@@

The algorithm is a variant of the ElGamal scheme. Its security is based on
the Discrete Logarithm Problem\index{logarithm problem!discrete}. The DSA public and private key and its
procedures for signature and verification are summarised in crypto procedure~\ref{dsasigproc}.

@@

\paragraph*{Signing}\strut\\
\begin{tabular}{l@{ }l}
$m$ & the message to be signed\\
$k$ & choose at random\index{random}, less than $q$\\
$r$ & $= (g^k \; \mbox{ mod } p) \mbox{ mod } q$\\
$s$ & $= (k^{-1}(\mbox{SHA-1}(m) + xr)) \mbox{ mod } q$
\end{tabular}

\emph{Remark:}
\begin{itemize}
\item $(s,r)$ is the signature.
\item The security of the signature depends not only on the mathematical
  properties, but also on using a good random source\index{random}  for $k$.
\item SHA-1 \index{SHA-1} is a 160-bit hash function.
\end{itemize}
\paragraph*{Verifying}\strut\\
\begin{tabular}{l@{ }l}
$w$ & $= s^{-1} \;  \mbox{ mod } q$\\
$u_1$ & $= (\mbox{SHA-1}(m)w) \mbox{ mod } q$\\
$u_2$ & $= (rw)  \mbox{ mod } q$\\
$v$ & $= (g^{u_1}y^{u_2}) \mbox{ mod } p)  \mbox{ mod } q$\\

@@

% --------------------------------------------------------------------------
\vskip + 15pt
\section{Public key certification}
\index{certification!public key}
\index{PKI}
The aim of public key certification is to guarantee the connection between a
public key and a user and to make it traceable for external parties. In cases in
which it is impossible to ensure that a public key really belongs to a
particular person, many protocols are no longer secure, even if the individual
cryptographic modules cannot be broken.

% --------------------------------------------------------------------------
\vskip + 15pt
\subsection{Impersonation attacks}\label{Impersonalisierungsattacke}
\index{impersonation attack}
Assume Charlie has two pairs of keys (PK1, SK1) and (PK2, SK2), where SK denotes
the secret key and PK the public key. Further assume that he manages to palm off
PK1 on Alice as Bob's public key and PK2 on Bob as Alice's public key (by
falsifying a public key directory).

@@

Another name for this type of attack is \index{attack!man-in-the-middle}
``man-in-the-middle attack''. Users are promised
protection against this type of attack by public-key
certification, which is intended to guarantee the
\index{authenticity} authenticity of public keys. The most common
certification method is the X.509 standard.

% --------------------------------------------------------------------------
\vskip + 15pt
\subsection{X.509 certificate}
\index{X.509} 
Each participant who wants to have an X.509 certificate (\cite{ds:X.509})
verifying that his public key belongs to a real person consults what is
known as a \index{certification authority (CA)} certification authority
(CA)\footnote{%
Often called trust center, if the certificates are not only offered
to a closed user group.
}. 
He proves his identity to this CA (for example by showing his ID). 
The CA then issues him an electronic document (certificate) which 
essentially contains the name of the certificate-holder and the name
of the CA, the certificate-holder's public key and the validity period
of the certificate. The CA then signs the certificate using its secret key.
